---
title: "Group Project 2"
author: "Prof. Ken Field"
date: "07 Mar 2023"
output:
  html_document:
    df_print: paged
subtitle: Biology 368/664 Bucknell University
---

## Objectives

1. Critically read a scientific paper and assess the evidence provided
2. Complete reporting checklists to determine if best practices were followed prior to publication
3. Assess a published paper for issues with reproducibility and transparency
4. Identify questionable research practices and describe appropriate methods to prevent them
5. Reproduce the analysis for one of the main figures of the paper (Homework 09)

```{r Load Libraries, include=FALSE}
if (!require("knitr")) install.packages("knitr"); library(knitr)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## Paper Assignment

Complete the survey by 11am on Thursday March 9th.

https://docs.google.com/forms/d/e/1FAIpQLSexRCMY3TyynwE4-OH-rbenLfs7j7AvJWnlgUwdFuYf7f5dqA/viewform?usp=sf_link

You will learn the paper that you are assigned in class on Thursday, Mar 9. 
Groups will be assigned within each Thursday section.

```{r}
papers <- dir("papers", pattern = "pdf", full.names = TRUE, ignore.case = TRUE)
papers <- gsub("papers/","",papers)
papers <- gsub(".pdf","",papers)
papers
```

## Groups

We will work in groups of 2 or 3 for this project, with each group being different from the Project 1 groups.

```{r echo=FALSE, results='asis'}
Project2 <- read_csv("../../00-Syllabus/roster_2023.csv") %>%
  arrange(NAME) %>%
  mutate(Paper = c(
    "[Guan.2021]",
    "[Pruitt.2016]",
    "[Le.2014]",
    "[Kern.2013]",
    "[Estruch.2013]",
    "[Sato.2013]",
    "[Guan.2021]",
    "[Abdallah.2020]",
    "[Pruitt.2016]",
    "[Woo.2004]",
    "Laskowski",
    "[Abdallah.2020]",
    "[Sato.2013]",
    "Laskowski",
    "Modlmeier",
    "[Le.2014]",
    "[Kern.2013]",
    "[Estruch.2013]",
    "[Woo.2004]",
    "Modlmeier"
  ))

table(Project2$Paper)
Project2 %>%
  arrange(Paper) %>%
  kable()
```

## Grading

Each student will be expected to complete the following tasks to earn 85% of the points available for this assignment (21/25).

- Read and discuss your paper with your group partners
  + Focus on the methods and results
  + For the major conclusion(s) evaluate the strength of the supporting evidence
- Complete the `nr-editorial-policy` and `nr-reporting-summary` checklists (*open in Adobe Acrobat*)
  + If the copy in the Repo doesn't work, download these forms at the links below and open in Acrobat
  + For any deficient items on the checklists, describe the information needed
  + Refer to nr-reporting-life-sciences-research.pdf for guidelines
- Determine if you could repeat the analysis that is most essential to conclusions of the paper
  + If data or methods are not available, look for follow-up studies that provide the required details
  + If follow-up studies are not available, describe the missing information needed to perform the analysis
- For each Questionable Research Practice discussed in Fraser *et al*:
  + Find evidence to exclude this QRP, or
  + Describe steps that should have been taken to prevent this QRP (see Forstmeier *et al*)
- Attempt to reproduce one of the key analyses from the paper
  + Evaluate the Data Accessibility for the manuscript, using the Nature Publishing Group Guidelines (link below)
  + For HW09 obtain the data from the paper and reproduce one of the analyses
- Present summary of reproducibility assessment in class

## Download Links
Reporting standards for Nature Journals: 
https://www.nature.com/nature-portfolio/editorial-policies/reporting-standards

Editorial Policy Checklist: 
https://www.nature.com/documents/nr-editorial-policy-checklist.pdf

Reporting Summary Checklist for Life Sciences Research:
https://www.nature.com/documents/nr-reporting-summary.pdf

Data availability:
https://www.nature.com/nature-portfolio/editorial-policies/reporting-standards#availability-of-data

## Acknowledgements

Each group should use an **Acknowledgements** section to document the participation of each member and the collaboration within groups.
Full credit for the assignment will require equal participation of each group member.
