---
title: "Lab 04"
author: "Biology Student"
date: "07 Feb 2023"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Libraries
Use the following syntax to load any libraries that you need. 
This code will prevent errors that will interfere with knitting of the Rmd file.
Also note that the `include=FALSE` option will prevent this code chunk from appearing in the markdown file.

We will always load the entire tidyverse library, which includes *dplyr*, *ggplot2*, *readr*, and other packages.

```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

As tidyverse loads, look to see the Conflicts that are noted. 
In particular, you will see that stats::filter() has been masked (replaced) by dplyr::filter(). 
If you just use the filter() command, it will now use dplyr::filter(). 
If you wanted to use the other version, you would have to call it explicitly like stats::filter(). 

## Objectives for Lab 4

1. Introduction to *dplyr* grammar and functions
2. Exploratory data analysis using *dplyr* 
3. Knit to produce a github-formatted output of both your code and the output

```{r dplyr help, include=FALSE}
?dplyr
# Run the line below to view detailed help about dplyr (but keep it commented to allow knitting)
#browseVignettes(package = "dplyr")
```

## Exploratory Data Analysis with R

This lab is adapted from Roger Peng's *Exploratory Data Analysis with R*, Chapter 4.
https://leanpub.com/exdata/read_full

## Verbs in dplyr

One of the main parts of the *dplyr* data grammar is the use of verbs to manipulate data frames.
There are six main verbs in *dplyr*:

- `filter()` to select cases based on their values.
- `arrange()` to reorder the cases.
- `select()` and `rename()` to select variables based on their names.
- `mutate()` and `transmute()` to add new variables that are functions of existing variables.
- `summarize()` to condense multiple values to a single value.
- `sample_n()` and `sample_frac()` to take random samples.

The preferred syntax of *dplyr* will use a data frame first, followed by a verb. 
The output will be another (altered) data frame. 
However, these verbs can also be used like traditional R functions where you have the verb(subject) syntax.
We will use this traditional syntax to learn what these functions do, and then introduce the preferred (pipeline) syntax.

## Loading data

For this exercise we will be using data from Prof. Reeder's fieldwork in Uganda on *Epomophorus labiatus*, little epauletted fruit bats. 
Epauletted fruit bats were collected at two different times of the year for this study and a variety of health metrics were measured.

Note that this data should not be shared outside this class without permission.

```{r}
BatData <- read_csv("../04-Data/UgandaBatsFilteredMetrics.csv", 
    col_types = cols(`Collection date` = col_date(format = "%m/%d/%y")))
# The col_date format was used to make sure that the Collection date was interpreted properly.
# See ?as.Date for more information and examples.

# Below is a very handy way to eliminate those annoying spaces in the column names!
names(BatData) <- make.names(names(BatData))

```

The make.names() function will save you lots of trouble if you use it before you begin any data exploration or analysis. 
Not only does it make the columns easier to refer to by removing the spaces (and no more backticks!), 
it also checks for other illegal characters in the column names. 
For example, did you know that column names can't start with a number? (I always forget that rule!)

Loading the data will give you an error if the read_csv() function can't find it at the location specified. 
You will have to change the path or else make sure that the file is located where it is looking.

After you load the data, check that it has loaded correctly by checking the structure of the data frame. 
The `str()` function is very useful for getting to know the structure of your data.

```{r}

```

## select()

The select() function can be used to select **columns** of a data frame that you want to focus on. Often you’ll have a large data frame containing “all” of the data, but any given analysis might only use a subset of variables or observations. The select() function allows you to get the few columns you might need.

Suppose we wanted to take the first 3 columns only. There are a few ways to do this. We could for example use numerical indices. But we can also use the names directly.

```{r}
names(BatData)[1:3]
BatData.subset <- dplyr::select(BatData, Field.number:Season)
head(BatData.subset)
```

Note that the : normally cannot be used with names or strings, but inside the select() function you can use it to specify a range of variable names.
You can also omit variables using the select() function by using the negative sign. With select() you can do
`select(BatData, -(Collection.date:Season))`

If you don't specify which select function to use and you have more than one package with a select function, then you will generate an error.
Thanks to the "conflicted" package, that error is informative and tells you how to fix it.
You can see that we specified the package to use by 'dplyr::select'. 

The select() function also allows a special syntax that allows you to specify variable names based on patterns. 
So, for example, if you wanted to keep every variable that ends with a “2”, we could use `subset <- select(chicago, ends_with("2"))`

Use select() to make a new BatData.subset that only contains the Field number, Season, Class, Forearm length (FA length), and Mass for each bat.

```{r}

```

## filter()

The filter() function is used to extract subsets of **rows** from a data frame. This function is similar to the existing subset() function in R but is quite a bit faster.

Suppose we wanted to extract the rows where the Mass of the bat was less than 55 g, which may indicate that the bat is a juvenile or subadult.

```{r}
BatData.filter <- filter(BatData, Mass < 55)
str(BatData.filter)
```

There are only 41 bats in this filtered list. Use summary() to compare the Masses of this group of bats to those of the entire set.

```{r}

```

You could also use filter to select based on more complex logical sequences using the & operator (or other logical operators). 
Compare the number of male bats that are under 55 g to the number of female bats that are under 55 g using the filter() function.

```{r}

```

## arrange()

The arrange() function is used to reorder rows of a data frame according to one of the variables/columns. Reordering rows of a data frame (while preserving corresponding order of other columns) is normally a pain to do in R. The arrange() function simplifies the process quite a bit.

Here we can order the rows of the data frame by date, so that the first row is the earliest (oldest) observation and the last row is the latest (most recent) observation.

```{r}
BatData <- arrange(BatData, Collection.date)
head(select(BatData, Field.number:Class))
tail(select(BatData, Field.number:Class))
```

Columns could also be arranged in descending order: `arrange(BatData, desc(Collection.date))`

## rename()

Renaming a variable in a data frame in R is surprisingly hard to do!
The rename() function is designed to make this process easier.
The first column of our data frame is the Id number that was given to each bat in the field.
The syntax inside the rename() function is to have the new name on the left-hand side of the = sign and the old name on the right-hand side.

```{r}
BatData <- rename(BatData, Id = Field.number)
head(select(BatData, Id:Class))
```

That's nicer!

## mutate()

The mutate() function exists to compute transformations of variables in a data frame.
It is one of the most useful dplyr verbs.
Often, you want to create new variables that are derived from existing variables and mutate() provides a clean interface for doing that.

For example, we can use Forearm length and Mass to calculate a body condition index (BCI) that will help us measure if a bat has a lower than usual Mass for its size. 

```{r}
BatData <- mutate(BatData, BCI = Mass / FA.length)
summary(BatData$BCI)
```

Use the other two lengths that were measured (total body length and hind foot length) to calculate alternative BCIs. 

```{r}

```

There is also the related transmute() function, which does the same thing as mutate() but then drops all the non-transformed variables.
We will use mutate() much more often than transmute().

## group_by()

The group_by() function is used to generate summary statistics from the data frame within categories defined by a variable.
For this dataset, the column *class* can be used to segregate the bats by sex, age, and reproductive status.
In conjunction with the group_by() function we often use the summarize() function.
The group_by() is VERY valuable for making statistical comparisons or visual comparisons between groups.

```{r}
BatClasses <- group_by(BatData, Class)
summarize(BatClasses, Mass=mean(Mass, na.rm = TRUE), BCI=mean(BCI, na.rm = TRUE))
```

## Pipelines

The output of one verb can be used by another verb in a **pipeline**. The pipeline is indicated by the pipe operator: `%>%`.
Pipelines can be used even if only a single verb is needed. 
For example, the code `filter(BatData, Mass < 55)` can be rewritten as: `BatData %>% filter(Mass < 55)` 

The pipeline operater %>% is very handy for stringing together multiple dplyr functions in a sequence of operations. 
Notice above that every time we wanted to apply more than one function, the sequence gets buried in a sequence of nested function calls that is difficult to read: 
`third(second(first(x)))`
This nesting is not a natural way to think about a sequence of operations. 
The %>% operator allows you to string operations in a left-to-right fashion:
`first(x) %>% second %>% third`
or even more clear:
`x %>% first %>% second %>% third`

Note that the pipeline operator is so important that it has now been made a part of base R.
As of R 4.1, the native operator |> can be used without loading any packages. 
If you have loaded the tidyverse package, like we always do in this class, then you can use either `%>%` or `|>`. 
The Rstudio shortcut Ctrl + Shift + M can be programmed to use either one, although the default is %>%.
Most of the time these two ways of piping data are equivalent, but there is one not uncommon situation where %>% is preferred. 
So if you try |> and it doesn't work, give %>% a try.

The example from the previous code chunk can be expressed more clearly using a pipeline:

```{r}
BatData %>% 
  group_by(Class) %>%
  summarize(Mass=mean(Mass, na.rm = TRUE), BCI=mean(BCI, na.rm = TRUE))
```

This way we don’t have to create a set of temporary variables along the way or create a massive nested sequence of function calls.
Notice in the above code that I pass the BatData data frame first to group_by(), but then afterwards I do not have to pass the first argument to summarize(). 
Once you travel down the pipeline with %>%, the first argument is taken to be the output of the previous element in the pipeline.

Use a pipeline to compare Mass and BCI of each of the Classes of bat *in each Season*. 
(Hint: group_by can accept more than one Name)

```{r}

```

## Summary

The dplyr package provides a concise set of operations for managing data frames. 
With these functions we can do a number of complex operations in just a few lines of code. 
In particular, we can often conduct the beginnings of an exploratory analysis with the powerful combination of group_by() and summarize().
Because it is also part of the tidyverse, you can also use pipelines to feed data to *ggplot2*!

## Knitting
Who knew that you were going to be spending time knitting in your Biology class? 
When you *knit* an Rmd file, you produce the output directed by the output: line in the header of the Rmd. 
For this class, that will typically be a github_document, but there are other times that a html_document or pdf_document might be more appropriate.

To knit your file, simply press the Knit button at the top of this panel. 
First, this saves the Rmd (so make sure you want to do that).
All of the code should then be run (similar to clicking Run All) and then a window should open displaying the .md file that was produced.
This github markdown file will need to be pushed to your repo for future assignments.

### Common problems with knitting
The single most common problem is when different code chunks have identical names. 
For whatever reason this is forbidden and RStudio is unable to handle this issue on its own. 
Either never name your code chunks (which isn't ideal for organizing your analysis), or always give them unique names. 
When the knit process fails it will tell you the name of the duplicate code chunk, so that you can track it down.

There are certain interactive functions in R that are incompatible with knitting. 
Because these functions require input from the user, they are not able to be knit into a static output. 
