---
title: "Lab 05 Warmup"
author: "Prof. Ken Field, Bucknell University"
date: "2/10/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## Loading Data

Scientists have been asked to investigate whether jelly beans cause acne. 
They enrolled 200 high school students in a study where the control group (n=100) avoided all jellybeans for 2 months.
The treatment group (n=100) were given a free bag of jellybeans every week and were told to consume all of them before the next week.
At the end of the study, the acne level of each participant was scored.

```{r}
acne <- read_csv("../04-Data/acne.csv")
str(acne)
acne <- acne %>% 
  mutate(Group = as.factor(Group)) %>%
  select(Score, Group)
str(acne)
```

## Statistical Test

The scientists determined that there was no link between jelly beans and acne.

```{r}
t.test(Score ~ Group, data = acne)
ggplot(acne, aes(x=Group, y=Score)) + 
  geom_boxplot()
```

But people are skeptical and think that it might not be all jellybeans, but only those of a certain color.
So the scientists stop playing Minecraft long enough to conduct a second study. 
This time they have a control group (n=10) and twenty groups (n=10 per group) that each eat bags of only a single color of jellybean.

```{r}
acne2 <- read_csv("../04-Data/acne2.csv")
acne2 <- acne2 %>% 
  select(Score, Color) %>%
  mutate(Color = as.factor(Color))
str(acne2)

# a for loop probably would have been better!
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "purple"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "brown"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "pink"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "blue"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "teal"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "salmon"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "red"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "turquoise"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "magenta"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "yellow"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "grey"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "tan"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "cyan"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "green"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "mauve"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "beige"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "lilac"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "black"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "peach"))
t.test(Score ~ Color, data = filter(acne2, Color == "control" | Color == "orange"))
```

## Whoa

```{r}
ggplot(filter(acne2, Color == "control" | Color == "green"), aes(x=Color, y=Score)) + 
  geom_boxplot()
```

https://xkcd.com/882/

## Linear Model Approach

Note that a linear model would produce different results. 
First, to set one of the levels as the reference (the control group), we need to
make sure it is the first level of the factor.

```{r}
jellybeans = c("control","purple","brown","pink","blue","teal",
               "salmon","red","turquoise","magenta","yellow",
               "grey","tan","cyan","mauve","beige",
               "green","lilac","black","peach","orange")
               
acne2 <- read_csv("../04-Data/acne2.csv")

acne3 <- acne2 %>%
  mutate(Color = factor(Color, levels = jellybeans))
colorlm <- lm(Score ~ Color, data = acne3)
summary(colorlm)
```

We find that because the linear model is testing the significance of each level of the factor
within the context of the whole dataset, that it produces different p values.
This is a better approach (if the assumptions of the linear model are met).

```{r}
plot(colorlm)
```

That doesn't look bad, but we can definitely see the limitations of having only
10 subjects per group.

## Multiple Testing Correction

We need to adjust the p-values for inflated false discoveries when conducting multiple tests.
The original, and most conservative method for multiple test correction was described by Bonferoni.

```{r}
color_pvalues <- coef(summary(colorlm))[,4]
p.adjust(color_pvalues, method = "bonferroni")
```

Acknowledgement: https://stackoverflow.com/questions/23838937/extract-pvalue-from-glm

## Behind the numbers

Both acne datasets were generated randomly using a normal distribution.
Note that if you run this code chunk and then rerun the analysis above,
you will get different results!

```{r}
acne <- data.frame(round(rnorm(200,10,5)))
names(acne) = c("Score")
acne$Group <- c(rep("control", each=100), rep("jellybeans", each=100))

jellybeans = c("control","purple","brown","pink","blue","teal",
               "salmon","red","turquoise","magenta","yellow",
               "grey","tan","cyan","mauve","beige",
               "green","lilac","black","peach","orange")

acne2 <- data.frame(round(rnorm(210,10,5)))
names(acne2) = c("Score")
acne2$Color <- c(rep(jellybeans, each=10))
```