---
title: "Lab 03"
output: github_document
author: "Ken Field"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("NHANES")) install.packages("NHANES"); library(NHANES)
if (!require("car")) install.packages("car"); library(car)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## NHANES Dataset

Data from the US National Health and Nutrition Examination Study
https://www.cdc.gov/nchs/nhanes/index.htm

This is survey data collected by the US National Center for Health Statistics (NCHS) which has
conducted a series of health and nutrition surveys since the early 1960’s. Since 1999 approximately
5,000 individuals of all ages are interviewed in their homes every year and complete the health
examination component of the survey. The health examination is conducted in a mobile examination
center (MEC).

A description of the variables can be found in the package description.

```{r NHANES}
data("NHANES")
summary(NHANES)
help(NHANES)
```

Several of the variables are categorical.
Let's convert them into factors all at once so they are handled by R properly.
We will learn a better way to do this in the future, but for now this will work.

```{r}
NHANES <- as.data.frame(unclass(NHANES), stringsAsFactors=TRUE)
```


## Model of Height Data

First look at the entire data set.

```{r Histogram}
ggplot(NHANES) +
  aes(x = Height) +
  geom_histogram(bins=100) +
  theme_cowplot()
```

Now lets just look at the data for children aged 2-15. 
Also drop any row that has NA for Height.

```{r Children Subset}
NHANES_child <- filter(NHANES, (Age >= 2 & Age <= 15))
NHANES_child <- drop_na(NHANES_child, Height)
ggplot(NHANES_child) +
  aes(x = Height) +
  geom_histogram(bins=100) +
  theme_cowplot()
```

## Model of the data

First, we will test to see if a simple model of using mean +- standard deviation
can predict the height of children. 

```{r Simple Model}
mean(NHANES_child$Height)
sd(NHANES_child$Height)
summary(NHANES_child$Height)
```

The error is large when using mean to predict the height. 
If this data had a normal distribution, this model would be better.

```{r}
error_mean <- NHANES_child$Height - mean(NHANES_child$Height)
ggplot(NULL, aes(error_mean)) +
  geom_histogram(bins = 100) + 
  xlim(-60, 60) +
  xlab("Error when predicting height with mean")
rmse_mean <- sqrt(mean(error_mean**2))
rmse_mean
```

More sophisticated models are necessary!

Note that below we are storing the plots as objects in memory so that we can plot
several at once. 

```{r Age}
p1 <- ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(position = "jitter",size=0.05) +
  scale_x_continuous(breaks = seq.int(0, 20, 2)) +
  ggtitle('A: original data') +
  theme_cowplot()

lmResultHeightOnly <- lm(Height ~ Age + 0, data=NHANES_child)
lmResultHeightOnly
summary(lmResultHeightOnly)
rmse_heightOnly <- sqrt(mean(lmResultHeightOnly$residuals**2))
rmse_heightOnly

p2 <-  ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(position = "jitter",size=0.05) +
  scale_x_continuous(breaks = seq.int(0, 20, 2)) + 
  annotate('segment',x=0,xend=max(NHANES_child$Age),
           y=0,yend=max(lmResultHeightOnly$fitted.values),
           color='blue',lwd=1) + 
  ggtitle('B: age') +
  theme_cowplot()

lmResultHeight <- lm(Height ~ Age, data=NHANES_child)
lmResultHeight
summary(lmResultHeight)
rmse_height <- sqrt(mean(lmResultHeight$residuals**2))
rmse_height

p3 <- ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(position = "jitter",size=0.05) +
  scale_x_continuous(breaks = seq.int(0, 20, 2)) + 
  geom_smooth(method='lm',se=FALSE) + 
  ggtitle('C: age + constant') +
  theme_cowplot()

plot_grid(p1,p2,p3,ncol=2)  ## Another great function of the cowplot package
```

This model is much better than using only the mean.

Can we make it even better?

```{r}
ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(aes(colour = Gender), 
             position = "jitter", 
             alpha = 0.8,
             size=0.05) +
  geom_smooth(method='lm',aes(group = Gender, 
                              colour = Gender)) + 
  theme(legend.position = c(0.25,0.8)) + 
  ggtitle('D: age + constant + gender') +
  theme_cowplot()

lmResultGenderAge <- lm(Height ~ Gender + Age, data=NHANES_child)
lmResultGenderAge
summary(lmResultGenderAge)
```

## Model Comparisons

We now want to know how much better the new model is than the one that didn't include gender.
The anova() function provides a convenient way to compare two nested models.
This is described very nicely here:
https://www.r-bloggers.com/2021/05/how-to-compare-nested-models-in-r/

```{r}
anova(lmResultHeight, lmResultGenderAge)
```
As you can see, the p value for the comparison is much less than 0.05, giving us high confidence that second model is a better fit for the data than the first model.


## Test Assumptions

Linear models make the following assumptions:
1. Independence of data points
2. Normally distributed residuals
3. Homoscedasticity

### Check for autocorrelation

The Durbin-Watson test checks for autocorrelation in a linear model.
The null hypothesis is no correlation among the residuals.
The function dubrinWatsonTest() is part of the car package.

```{r}
durbinWatsonTest(lmResultGenderAge)
```

Since this p-value is less than 0.05, we can reject the null hypothesis and conclude that the residuals in this regression model are autocorrelated.

This does not mean that we can't use a linear modeling approach. 
But we do have to be careful about interpreting the results. 
In THIS CASE, the autocorrelation is presumably due to the fact that tall 4 year-olds become tall 5 year-olds, etc. 
So it is not surprising in a set of data that have a time component to find autocorrelation.

### Residual distribution

We need to check that the residuals are similar at each age and for both genders by looking at plot 3 below.

```{r}
plot(lmResultGenderAge)
```

That looks good. 
Gender does not make a consistent difference. 
There is a slight increase in the standardized residuals at higher ages, but it is not very large.


Linear models assume that the **residuals** have a normal distribution.
By using the plot() function on the linear model we can see the distribution
of the residuals in the first plot.

```{r}
plot(lmResultGenderAge)
```

In the future we will learn a better way to confirm the distribution of the residuals, 
but for now just look at the first two plots to see how they look. 

### Normally distributed residuals

Check residual distribution: 
```{r}
hist(lmResultGenderAge$residuals, breaks = 20)
```

Well that looks pretty good. 
And when we check the QQ Plot (plot 2, above): 
We find that most of the points lie along the reference line.

### Homoscedasticity

See Residuals vs Fitted plot (plot 1, above): 
The line is pretty flat but not perfect. 
The residuals are not randomly distributed but seem to increase with the fitted values. 
There is slight evidence of kurtosis (S-shaped pattern), but not too bad.

Overall, this model is showing that it is a fairly good fit for the assumptions.

It may be possible to improve the model using additional variables...

## Acknowledgements
=========================
Linear regression assumptions: 
https://rpubs.com/aryn999/LinearRegressionAssumptionsAndDiagnosticsInR 
https://www.r-bloggers.com/2016/01/how-to-detect-heteroscedasticity-and-rectify-it/ 
https://web.ma.utexas.edu/users/mks/statmistakes/modelcheckingplots.html
https://bookdown.org/jimr1603/Intermediate_R_-_R_for_Survey_Analysis/testing-regression-assumptions.html
