---
title: "Homework 07"
author: "Biology Student"
date: "09 Feb 2023"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## Objectives

For this assignment you will need to identify a dataset from a public data archive use it to conduct the following steps:

- [ ] Data Wrangling
- [ ] Data Exploration
- [ ] Data Visualization
- [ ] Hypothesis testing using a generalized linear model

Your first step is to identify the dataset that you want to use and confirm that it is suitable for this assignment.

1. Contains between 100 and 10,000 observations.
2. Has a quantitative outcome variable (not logistic like survival).
3. Measures 4 or more potential explanatory variables (quantitative or categorical).
4. Is NOT a time series.

## Public Data Archives

You can find your dataset either by browsing a data archive or by finding a paper that includes the raw data as supplementary material (or links to a data archive).
There are many specialized data archives (http://oad.simmons.edu/oadwiki/Data_repositories), but these are some general ones that include biology:

- Data Dryad: https://datadryad.org/search
- Open Science Framework: https://osf.io/
- Figshare: https://figshare.com/
- Mendeley Data: https://data.mendeley.com/
- TreeBASE: https://www.treebase.org
- Zenodo: https://zenodo.org/
- Dataverse: https://dataverse.harvard.edu/

Spend no more than *one hour* browsing those data archives or papers looking for a dataset to use.
If you are unable to find one that is suitable, you may use the built in dataset called gapminder. 

```{r Load Gapminder}
if (!require("gapminder")) install.packages("gapminder"); library(gapminder)
data(gapminder)
glimpse(gapminder_unfiltered)
```

That only represents a tiny fraction of the gapminder data archive. 
To access the other data, go to https://www.gapminder.org/data/ and find at least three different datasets to download and analyze (after joining them). 
For your analysis you will need to have one dependent variable and at least two potential explanatory variables.

