---
title: "Homework 07"
author: "Biology Student"
date: "09 Feb 2023"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## Objectives

For this assignment you will need to identify a dataset from a public data archive use it to conduct the following steps:

- [ ] Not completed
- [x] Completed

1. Data Wrangling
- [ ] Dataset meets the criteria described below (at least before cleaning begins)
- [ ] Imported data checked to verify that it matches the dimensions described in the paper or help file
- [ ] Structure of dataframe checked to verify that each variable has proper type (i.e. factors, numeric, etc.)
- [ ] Some method used to look at data to ensure that it is correct (glimpse, headtail, etc.)
2. Data Exploration
- [ ] An appropriate summary is generated of the outcome variable (mean, median, summary, or other)
- [ ] Each potential explanatory variable is summarized (could be as simple as skim())
- [ ] The distribution of numeric variables are shown (histograms and/or QQ plots)
3. Data Visualization
- [ ] The outcome variable is plotted against one or more of the explanatory variables
- [ ] An appropriate visualization is used for the type of data (boxplot, scatter plot, or histogram)
- [ ] If some of the explanatory variables are quantitative was the correlation between them and the outcome variable checked? 
- [ ] The visualization(s) use best principles for showing all of the data 
- [ ] The visualization(s) demonstrate the major trends or contrasts in the data and help lead to the hypothesis
- [ ] Results of the exploration and visualization are interpreted clearly to justify hypothesis
4. Hypothesis testing using a generalized linear model
- [ ] An appropriate glm model is used (linear model or negative binomial)
- [ ] Model selection is used to derive a "best" model (backward selection using StepAIC) 
- [ ] Model visualization is performed to show the outcome of the model (interact_plot)
- [ ] Model is interpreted using AIC, estimates (coefficients), and p-values appropriately

Your first step is to identify the dataset that you want to use and confirm that it is suitable for this assignment.

- [ ] Contains between 100 and 10,000 observations.
- [ ] Has a quantitative outcome variable (not logistic like survival).
- [ ] Measures 4 or more potential explanatory variables (quantitative or categorical).
- [ ] Is NOT a time series.

## Public Data Archives

You can find your dataset either by browsing a data archive or by finding a paper that includes the raw data as supplementary material (or links to a data archive).
There are many specialized data archives (http://oad.simmons.edu/oadwiki/Data_repositories), but these are some general ones that include biology:

- Data Dryad: https://datadryad.org/search
- Open Science Framework: https://osf.io/
- Figshare: https://figshare.com/
- Mendeley Data: https://data.mendeley.com/
- TreeBASE: https://www.treebase.org
- Zenodo: https://zenodo.org/
- Dataverse: https://dataverse.harvard.edu/

Spend no more than *one hour* browsing those data archives or papers looking for a dataset to use.
If you are unable to find one that is suitable, you may use the built in dataset called gapminder. 

```{r Load Gapminder}
if (!require("gapminder")) install.packages("gapminder"); library(gapminder)
data(gapminder)
glimpse(gapminder_unfiltered)
```

That only represents a tiny fraction of the gapminder data archive. 
To access the other data, go to https://www.gapminder.org/data/ and find at least three different datasets to download and analyze (after joining them). 
For your analysis you will need to have one dependent variable and at least two potential explanatory variables.

