---
title: "Homework 05"
author: "Biology Student"
date: "09 Feb 2023"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Due Friday 17 Feb 2023 at 11:59pm

## Loading Libraries
Use the following syntax to load any libraries that you need. This code will prevent errors that will interfere with knitting of the Rmd file.
Also note that the `include=FALSE` option will prevent this code chunk from appearing in the markdown file.

We will always load the entire tidyverse library, which includes *dplyr*, *ggplot2*, *readr*, and other packages.

```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

*Exploratory Data Analysis Checklist*

_____________________________________

Follow Chapter 5 from *Exploratory Data Analysis with R* (https://leanpub.com/exdata/read_full) by Roger D. Peng to explore the Uganda Bat Dataset from Lab 4

You also may wish to follow along with the analysis in EDA5. 
To do that, you will need to access the data as shown below.

Remember to edit the Rmd to remove any irrelevant code for your assignment.

For each of the "Tasks" in this homework you will only need one or two code chunks and short explanations of your rationale and results. 
Some of them are very simple -- don't overthink it.

## Formulate your question

Always formulate your question(s) *before* you begin analyzing the data. 
(But you may look at the names of the variables measured to see what questions are possible.)

The data was collected from 103 bats in 2017. 
There are four potential dependent variables with units of mm: "Total Length", "Hind Foot", "Ear", and "FA length".
The potential dependent variable "Mass" is in g.

There are six potential explanatory variables: 
 - "Collection date"
 - "Season": Whether the bat was captured in the Rainy or Dry season
 - "Sex"                
 - "Class": Includes Sex, Age (sexual maturity), and pregnancy status
 - "Elevation": The elevation of the capture site. Note that it covers a small portion of the total habitat range for this species.
 - "Habitat description": Field notes on the environment of the capture site.

## Read in your data

If you want to follow along with the EDA5 reading:

```{r Load 2014 Ozone Data}
download.file("https://aqs.epa.gov/aqsweb/airdata/hourly_44201_2014.zip", "hourly_44201_2014.zip")
OzoneData <- read_csv("hourly_44201_2014.zip", col_types = "ccccinnccccccncnncccccc")
file.remove("hourly_44201_2014.zip")
```

Some functions like read_csv() can unzip files!

Note, if that download times out (takes more than 60 seconds), then you will have to download the file manually.

The Ozone data that is used in EDA can be found at the EPA website:
https://aqs.epa.gov/aqsweb/airdata/download_files.html

For the homework, you will need to use the bat data.

```{r Load Bat Data}
BatData <- read_csv("UgandaBatsFilteredMetrics.csv", 
    col_types = cols(`Collection date` = col_date(format = "%m/%d/%y")))
```

Remember to delete any unnecessary code chunk(s).

## Check the packaging

Validate that the data file is complete and properly loaded.
See above for the expectations.

## Run str()

I prefer the `glimpse()` function of dplyr over `str()`

Fix any problems that you find, such as converting strings to factors when appropriate.

Keep all of the data together, even if you do not plan on using some of the columns.

## Look at the top and the bottom of your data

## Check your “n”s

## Validate with at least one external data source

For the bat dataset, the information here will be helpful: https://eol.org/pages/327340/data

The species is *Epomophorus labiatus*, the Little Epauletted Fruit Bat. 

## Try the easy solution first

## Challenge your solution

This step can be skipped because we don't really know enough about bat data to derive a good "challenge".

## Follow up

Anser the questions in the follow-up section.

*Analytic Graphics*

___________________


Follow Section 6.1 and 6.2 to visualize the comparisons that were essential to answering the question.

## Show comparisons

This may be somewhat redundant with your work above.

## Show causality, mechanism, explanation, OR systematic structure (for 100% credit)

Note that we can't test **causality** in this dataset, but you can try to get a good handle on the variable(s) that show the best **correlation** with the dependent variable that you chose.
