---
title: "Project 3 Summary"
authors: "Zhao, Jiawen, Amano, Aiko, Brendel, Meredith T., Buabeng, Isaac K., DeSana, Andrew,
Douek, Naomi R., Frattasio, Molly O., Gao, Jasmine, Guzman, Dariel, Halbreiner, Claudia J., 
Ling, Tianze, Martinson, Katy A., Paula-Gil, Heather, Pring, Sam S., Prohofsky, Anna, 
Saum, Paul B., Smith, Meghan C., Sporea, Gloria A., Thrall, Abigail E., Umeh Jr., Micah N.
Zheng, Carol, Field, Ken A."
date: "13 Apr 2023"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
if (!require("knitr")) install.packages("knitr"); library(knitr)
if (!require("UpSetR")) install.packages("UpSetR"); library(UpSetR)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse) # Load tidyverse last
```

## Project 3 Groups

| NAME                   | EMAL                | Project3 | Tissue | Comparison |
|------------------------|---------------------|:--------:|:------:|:----------:|
| Buabeng, Isaac K.      | ikb001@bucknell.edu | 1        |  AMY   | Isolation  |
| Amano, Aiko            | aa037@bucknell.edu  | 2        |  AMY   | Treatment  |
| DeSana, Andrew         | ad049@bucknell.edu  | 3        |  BNST  | Isolation  |
| Prohofsky, Anna        | ap048@bucknell.edu  | 4        |  BNST  | Treatment  |
| Halbreiner, Claudia J. | cjh030@bucknell.edu | 5        |  FC    | Isolation  |
| Guzman, Dariel         | dg030@bucknell.edu  | 5        |        |            |
| Frattasio, Molly O.    | mof003@bucknell.edu | 6        |  FC    | Treatment  |
| Pring, Sam S.          | ssp008@bucknell.edu | 6        |        |            |
| Brendel, Meredith T.   | mtb018@bucknell.edu | 7        |  HIPP  | Isolation  |
| Thrall, Abigail E.     | aet009@bucknell.edu | 7        |        |            |
| Paula-Gil, Heather     | hpg004@bucknell.edu | 8        |  HIPP  | Treatment  |
| Douek, Naomi R.        | nrd005@bucknell.edu | 8        |        |            |
| Gao, Jasmine           | mg060@bucknell.edu  | 9        |  PVN   | Isolation  |
| Sporea, Gloria A.      | gas020@bucknell.edu | 9        |        |            |
| Umeh Jr., Micah N.     | mnu001@bucknell.edu | 10       |  PVN   | Treatment  |
| Saum, Paul B.          | pbs018@bucknell.edu | 10       |        |            |
| Smith, Meghan C.       | mcs023@bucknell.edu | 11       | HIPP-FC|   Tissue   |
| Ling, Tianze           | tl016@bucknell.edu  | 11       |        | Pair-Untr  |
| Martinson, Katy A.     | kam071@bucknell.edu | 12       |AMY-BNST|   Tissue   |
| Zheng, Carol           | yz017@bucknell.edu  | 12       |        | Pair-Untr  |


```{r Loading Data, message = FALSE}
# Isolation Comparison
isofilelist <- list.files(path = "voles/IsolationResults", pattern = "*.csv", full.names = T)
isofilenames <- list.files(path = "voles/IsolationResults", pattern = "*.csv", full.names = F)
isofiles <- lapply(isofilelist, read_csv)
isofilenames <- gsub(".csv", "", isofilenames)
names(isofiles) <- isofilenames
allisoresults <- bind_rows(isofiles, .id = "column_label")

# Treatment Comparison
treatfilelist <- list.files(path = "voles/TreatmentResults", pattern = "*.csv", full.names = T)
treatfilenames <- list.files(path = "voles/TreatmentResults", pattern = "*.csv", full.names = F)
treatfiles <- lapply(treatfilelist, read_csv)
treatfilenames <- gsub(".csv", "", treatfilenames)
names(treatfiles) <- treatfilenames
alltreatresults <- bind_rows(treatfiles, .id = "column_label")

# Tissue Comparison
tissuefilelist <- list.files(path = "voles/TissueResults", pattern = "*.csv", full.names = T)
tissuefilenames <- list.files(path = "voles/TissueResults", pattern = "*.csv", full.names = F)
tissuefiles <- lapply(tissuefilelist, read_csv)
tissuefilenames <- gsub(".csv", "", tissuefilenames)
names(tissuefiles) <- tissuefilenames
alltissueresults <- bind_rows(tissuefiles, .id = "column_label")
```

Note that `message = FALSE` was set to avoid outputting all of the column specifications.
Remove that option if you need to debug the column types.

### Isolation Comparisons

```{r}
# Verify that each results table has expected padj and fold change cutoffs
allisoresults %>%
  group_by(column_label) %>%
  summarise(max_padj = max(padj), min_foldchange = min(foldChange), 
            max_foldchange = max(foldChange))

# Count the number of genes differentially expressed
allisoresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("DOWN", ignore_case = T))) %>%
  summarize(Down = length(ID)) -> iso_summary_down
allisoresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("UP", ignore_case = T))) %>%
  summarize(Up = length(ID)) -> iso_summary_up

#Combine
iso_summary_down$column_label <- gsub(".DOWN", "", 
                                      iso_summary_down$column_label, ignore.case = TRUE)
iso_summary_up$column_label <- gsub(".UP", "", 
                                    iso_summary_up$column_label, ignore.case = TRUE)
iso_summary <- merge(iso_summary_up, iso_summary_down, by = "column_label")
iso_summary <- dplyr::rename(iso_summary, Comparison = column_label)
iso_summary
```

### Treatment Comparisons

Repeat the above summary for the treatment-dependent differential expression.

```{r}
# Verify that each results table has expected padj and fold change cutoffs
alltreatresults %>%
  group_by(column_label) %>%
  summarise(max_padj = max(padj), min_foldchange = min(foldChange), 
            max_foldchange = max(foldChange))

# Count the number of genes differentially expressed
alltreatresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("DOWN", ignore_case = T))) %>%
  summarize(Down = length(ID)) -> treat_summary_down
alltreatresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("UP", ignore_case = T))) %>%
  summarize(Up = length(ID)) -> treat_summary_up

#Combine
treat_summary_down$column_label <- gsub(".DOWN", "", 
                                      treat_summary_down$column_label, ignore.case = TRUE)
treat_summary_up$column_label <- gsub(".UP", "", 
                                    treat_summary_up$column_label, ignore.case = TRUE)
treat_summary <- merge(treat_summary_up, treat_summary_down, by = "column_label")
treat_summary <- dplyr::rename(treat_summary, Compartreatn = column_label)
treat_summary
```


### Tissue Comparisons

Repeat the above summary for the tissue-dependent differential expression.

```{r}
# Verify that each results table has expected padj and fold change cutoffs
alltissueresults %>%
  group_by(column_label) %>%
  summarise(max_padj = max(padj), min_foldchange = min(foldChange), 
            max_foldchange = max(foldChange))

# Count the number of genes differentially expressed
alltissueresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("DOWN", ignore_case = T))) %>%
  summarize(Down = length(ID)) -> tissue_summary_down
alltissueresults %>%
  group_by(column_label) %>%
  dplyr::filter(str_detect(column_label, regex("UP", ignore_case = T))) %>%
  summarize(Up = length(ID)) -> tissue_summary_up

#Combine
tissue_summary_down$column_label <- gsub(".DOWN", "", 
                                      tissue_summary_down$column_label, ignore.case = TRUE)
tissue_summary_up$column_label <- gsub(".UP", "", 
                                    tissue_summary_up$column_label, ignore.case = TRUE)
tissue_summary <- merge(tissue_summary_up, tissue_summary_down, by = "column_label")
tissue_summary <- dplyr::rename(tissue_summary, Compartissuen = column_label)
tissue_summary
```

Which manipulation (tissue, isolation, or treatment) resulted in the greatest numbers of genes meeting the padj and fc cutoffs?

Which tissue had the greatest numbers of differentially expressed genes in each comparison group?

## Gene-by-gene comparison

We would like to know if the same genes are affected by isolation and oxytocin treatment.

Compare the transcripts or genes identified as differentially expressed in your comparison to the other one in the same tissue.

We could use a Venn diagram to make this comparison, but like pie charts these are not always the best visualization. Or are they?
https://www.promusa.org/blogpost174-The-best-genomics-Venn-diagram-ever-deconstructed

The UpSetR package includes better tools for visualizing the intersections of multiple sets. 
https://cran.rstudio.com/web/packages/UpSetR/vignettes/basic.usage.html

The following code plots the intersection between the Isolation results for all of the files loaded. 

```{r}
de_batch <- allisoresults %>% 
  dplyr::select(Tissue = column_label, ID)

intersection <- de_batch %>% mutate(value =1) %>% spread(Tissue, value, fill =0 )
head(intersection)
intersection <- as.data.frame(intersection)
upset(intersection, nsets = 10)
upset(intersection, empty.intersections = "on", order.by = "freq", nsets = 10)
```

Use this example to compare your DE results to the other comparison in the same tissue.

```{r}
de_HIPP_iso <- allisoresults %>% 
  select(Tissue = column_label, ID) %>%
  filter(str_detect(Tissue, 'HIPP'))
de_HIPP_treat <- alltreatresults %>% 
  select(Tissue = column_label, ID) %>%
  filter(str_detect(Tissue, 'HIPP'))
de_HIPP <- bind_rows(de_HIPP_iso, de_HIPP_treat)

intersection_HIPP <- de_HIPP %>% mutate(value =1) %>% spread(Tissue, value, fill =0 )
head(intersection_HIPP)
intersection_HIPP <- as.data.frame(intersection_HIPP)
upset(intersection_HIPP, nsets = 10)
```

Were any of the same genes found differentially expressed in both comparisons? 

Yes, 7 genes were Up in treatment and Down in isolation and 2 genes were Up in both.

If so, which genes and were they regulated in the same direction in both?

```{r}
# Genes Up in both
intersection_HIPP %>%
  filter((HIPP.Isolation.Up == 1) & (HIPP.Treatment.Up == 1))
# Genes Up in treatment and Down in isolation
intersection_HIPP %>%
  filter((HIPP.Isolation.Down == 1) & (HIPP.Treatment.Up == 1))
```


Prediction:


## Acknowledgements

Using lapply() to run functions over a list: https://astrostatistics.psu.edu/su07/R/html/base/html/lapply.html

Converting a list into a single data frame: https://stackoverflow.com/questions/2851327/convert-a-list-of-data-frames-into-one-data-frame

UpSetR: https://cran.rstudio.com/web/packages/UpSetR/

Converting gene lists into UpSetR format: https://crazyhottommy.blogspot.com/2016/01/upset-plot-for-overlapping-chip-seq.html

Upset plots for genome analysis: https://genomespot.blogspot.com/2017/09/upset-plots-as-replacement-to-venn.html 

The best genomics Venn diagram ever?
https://www.promusa.org/blogpost174-The-best-genomics-Venn-diagram-ever-deconstructed

Reminder about how to bind rows:
https://dplyr.tidyverse.org/reference/bind.html