---
title: "Hibernation Dataset"
author: "Ken Field"
date: "24 Jan 2023"
output: html_document
---

This code helps set up the R markdown document to make a nice clean html file for sharing. 
Click the green triangle to run the code chunk. 

Note that all of your packages should be installed in the first code chunk.
Tidyverse should generally be the last package installed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("readxl")) install.packages("readxl"); library(readxl)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

For Homework 01 (only) you are *not* permitted to load any additional packages for your data analysis.

The dataset that you will be analyzing came from the following paper:
Heldstab SA, Isler K, van Schaik CP (2018) Hibernation constrains brain size evolution in mammals. Journal of Evolutionary Biology 31(10): 1582-1588. https://doi.org/10.1111/jeb.13353

The dataset was downloaded from Dryad https://datadryad.org/, a data repository for biology. The citation for this dataset:
Heldstab SA, Isler K, van Schaik CP (2018) Data from: Hibernation constrains brain size evolution in mammals. Dryad Digital Repository. https://doi.org/10.5061/dryad.753d06g

Download the xls data from that page or from the class github repository and make sure that it is in the same folder as this Rmd file.

```{r Import Data}
Supplementary_material <- read_excel("Supplementary material.xlsx", 
    col_types = c("text", "text", "numeric", 
        "text", "numeric", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "numeric", "text"))
# I have made sure that the columns of numbers that represent categories 
# are read in as characters instead of numbers
```

The description of the data:
List of species and data used for this study (1104 mammalian species) including references for brain mass, body mass, hibernation, diurnality, substrate use, diet and mid-latitude of geographical distribution data

```{r}
head(Supplementary_material)
summary(Supplementary_material)
names(Supplementary_material)
# Column names with spaces in them are annoying!
# To use them in R, you will need to surround them with backticks: `
summary(Supplementary_material$`brain mass`)
# The categories can be better interpretted as factors instead of characters:
summary(as.factor(Supplementary_material$diurnality))
```

Before you proceed with analyzing this dataset, check that it is complete and that you understand what each variable (column) means. You may need to refer to the paper (available in the repository).
![Heldstab et al.](Heldstab_et_al-2018-Journal_of_Evolutionary_Biology.pdf)

From the Methods section:

Hibernation was coded as a binary variable with (0) for nonhibernating and (1) for hibernating species with data from the published literature.

To control for diet quality, species were divided into four categories based on their main diet: herbivore or folivore (1), frugivore/folivore or granivore (2), frugivore/faunivore or omnivore (3) and faunivore, piscivore, carnivore or insectivore (4). A binary coding was used for activity period, with (0) for nocturnal, cathemeral or crepuscular species and (1) for diurnal species. For substrate use, each species was assigned to one of five substrate use categories: semi-aquatic (1), fossorial or semi-fossorial (2), terrestrial or semi-arboreal (3), arboreal (4) and volant (5).

After becoming familiar with the dataset, conduct exploratory data analysis of the key variables to test the hypothesis provided in the title of the paper: "Hibernation constrains brain size evolution in mammals". 
While you are exploring the data, come up with an alternative hypothesis and test it.

Completion of the assignment requires:
(1) sufficient exploration of the data to support your chosen method of analysis
(2) a graph that provides a clear illustration of support (or lack of support) for the paper's hypothesis
(3) a statistical test that quantifies the level of support for the paper's hypothesis

To receive full credit for the assignment you will need to provide: 
(1) step-by-step explanation of your exploration of the data to support your chosen method of analysis
(2) graphs that provide a clear illustration of support (or lack of support) for both hypothesis
(3) a statistical test that quantifies the level of support for your hypothesis

```{r}
print(Supplementary_material)
```

That was just a simple change.

## Acknowledgements
All assignments and exams should include an acknowledgements section where you cite all sources of help that you received for that project.
You DO NOT need to acknowledge help from Prof. Field or any TAs for the class.
You DO need to include links to any websites that you used, name any peers that you worked with, and include any prompts that you provided to OpenAI or other machine learning tools.